Testing Traversability of Protein Tunnels with Flexible Ligands Using Motion Planning

### 2017-06-15 paper.tex
scripts for making tables, experiments

\vskip -50pt
\abstract{
Understanding of interactions between enzymes and ligands is very important in enzymology, protein engineering or drug design. 
Molecular tunnels play an important role in catalytic and thermodynamic properties and they connect catalytic active sites with the solvent. 
Therefore, the geometrical and physico-chemical properties of tunnels need to be determined to understand the transport of ligand from and into the protein active site. 
Current methods describing the ligand passing through the tunnel are based either solely on the geometry of the tunnel or on molecular dynamic simulations. 
In this paper, we propose to analyze the ligand passing using sampling-based motion planning, namely using Rapidly Exploring Random Tree (RRT).
The tunnel is represented by a set of spheres, that are used to guide the growth of the RRT configuration tree in the related configuration space.
Multiple trajectories are computed using the proposed planner in order to analyze expected behavior of the ligand in the tunnel.
We introduce several approaches to visualization of detected tunnels which depict valuable information about various properties and behavior of the ligand.
Our proposed detection method, compared to the traditionally used geometrical analysis, provides more insight into the ligand possibility to pass through the tunnel. 
On the other hand, it is much less computationally demanding than methods based on molecular dynamic simulations. 
The method can be used to improve virtual screening results which are usually based only on determination of ligand binding in the active site but no information about ligand transport through the tunnel is provided.
}

@@

\begin{figure}[t]
\centering
{\footnotesize
\renewcommand{\arraystretch}{0.1}
\renewcommand{\tabcolsep}{0pt}
\begin{tabular}{ccccc}
\includegraphics[width=0.15\textwidth]{fig/motiv1} &
\includegraphics[width=0.17\textwidth]{fig/motiv2lab} &
\includegraphics[width=0.16\textwidth]{fig/motiv3} &
\hbox{
\vbox{
\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-1} }
\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-433}}
} 
}
\\
Protein 1CQW & Active site & Detected tunnels & Example of a  \\
             &            & (orange)          & trajectory
\end{tabular}
}
\caption{\label{fig::motiv}
    Example of tunnel detection in haloalkane dehalogenase.
}
\end{figure}

Understanding of interactions between proteins and other small molecules is crucial in many research fields, including drug design and protein engineering. 
These interactions are highly influence by the ability to transport the small molecule, called ligand, to the protein active site.
The active site can be considered as a deeply buried inner cavity with the ability to interact with the incoming ligand.
To get the ligand to the active site, there has to be the transportation path connecting the protein outer environment with the active site.
This paths, called tunnels, have to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand.

@@

The widely used tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3} or 
a 3D grid~\cite{sehnal2013mole,petrek2006caver}.
These methods compute the tunnels assuming a spherical ligand (probe), and the resulting tunnels are represented as a sequence of spheres.
Bio-chemists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and bottleneck (i.e.,
        the radius of the smallest sphere that can traverse the whole tunnel).
Typical ligands are however non-spherical and it is therefore difficult to estimate their traversability in tunnels computed for a spherical probe.
The decision based only on the spherical tunnels requires previous expertise in the domain and yet it may be imprecise.

@@

\red{presunu jinam}
This brings two important advantages: the volume of the configuration space to be searched is decreased, as not the whole
protein/ligand complex, but only its subspace need to be searched.
Consequently, the relative volume of the corresponding free region of the configuration space is increased, which leads to easier motion planning.
The proposed approach allows the researches to analyze a given tunnel for various ligands without necessity to compute the tunnel
again for each ligand.

@@

To guide tree growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of a tunnel sphere.
The random samples $\qrand$ are generated with probability $1-\gb$ from $\C$, otherwise they are generated
around the virtual goal sphere $v$.
%around which the random samples are generated with probability $\gb$.
%The random samples are not generated only from whole $\C$, as in the original RRT, but the sampling
%is guided using moving virtual goal~\cite{vonasek2009rrt}.
After the tree reaches the sphere $t_v \in T$, i.e., the distance of the tree to the center of $t_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are
not traversable or reachable by the ligand.
This is necessary in the dense protein structures, where it is not always possible to exactly follow the tunnel.
The algorithm terminates after the predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
from $N(t_v,\Sigma)$, where $\Sigma$ is diagonal matrix with diagonal entries equal to a parameter $\rv$, and the rotational
part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
The other two parameters (ligand index $l$ and scale $s$) of $\qrand$ can be left zero, as these are not used in the employed
metric for nearest-neighbor search.

@@

The result of the proposed planning method is a tree of collision-free configurations, from which a path leading from $\qinit$
to the nearest node towards $\qgoal$ can be found.
Due to stochastic nature of RRT, it is not guaranteed that the tree always reached the goal configuration $\qgoal$.
Despite, a trajectory leading as close as possible towards $\qgoal$ is extracted from the tree.

@@

The tested ligands are bigger than the tunnel with, therefore the ligands have been scaled down, and four
different scales were used: $\smin=\{0.3,0.4,0.5,0.6\}$.
For each of this scale, 50 collision-free initial states were found in the $2$\AA radius of the first sphere of the tunnel.
For each ligand, each minimal scale $\smin$ and each start configuration, 100 trajectories were computed using the proposed planner.
The parameters of the planner are: $\Imax=10000$, $m=50$, $\gb=0.9$, $\dt=1.5$\AA.

The average runtimes and sizes of the build tree are shown in Tab.~\ref{tab::rrt}.
For each trajectory

@@

\begin{table}
\centering
\caption{\label{tab::rrt}
    Runtime and sucess ratio of the planner
}
\input table.003.tex
\hskip 10pt
\input table.004.tex
\end{table}

@@

\section{Discussion}

guiding, parameters, different sampling, flexibility

The configuration tree can be expanded by up to $|\L|$ new configurations, as for each ligand conformation at most
one new configuration can be added.
The most time consuming part of the method is the collision-detection, which is called $m|\L|$ in each expansion step.
The collision detection can be evaluated fast using hierarchical methods like OBB trees, which is implemented e.g. in the OZCollide library~\cite{ozcollide}

>>>>

\vskip -50pt
\abstract{
Understanding of interactions between enzymes and ligands is very important in enzymology, protein engineering or drug design. 
Molecular tunnels play an important role in catalytic and thermodynamic properties and they connect catalytic active sites with the solvent. 
Therefore, the geometrical and physico-chemical properties of tunnels need to be determined to understand the transport of ligand from and into the protein active site. 
Current methods describing the ligand passing through the tunnel are based either solely on the geometry of the tunnel or on molecular dynamic (MD) simulations.
Although the tunnels can be computed fast using e.g., Voronoi diagrams, they cannot provide information about ligand traversability, as it is 
not included in the VD-based computation.
On the other side, MD simulations are too slow for a daily routine.
In this paper, we propose to analyze the ligand passing using sampling-based motion planning, namely using Rapidly Exploring Random Tree (RRT).
The tunnel is represented by a set of spheres, that are used to guide the exploration in the search space.
Multiple trajectories are computed for the ligand being tested.
We introduce several approaches for visualization of trajectories which depict valuable information about various properties and behavior of the ligand.
Our proposed detection method, compared to the traditionally used geometrical analysis, provides more insight into the ligand possibility to pass through the tunnel. 
On the other hand, it is much less computationally demanding than methods based on MD simulations.
The method can be used to improve virtual screening results which are usually based only on determination of ligand binding in the active site but no information about ligand transport through the tunnel is provided.
}

@@

\begin{figure}[b]
\centering
{\footnotesize
\renewcommand{\arraystretch}{0.1}
\renewcommand{\tabcolsep}{0pt}
\begin{tabular}{ccccc}
\includegraphics[width=0.15\textwidth]{fig/motiv1} &
\includegraphics[width=0.17\textwidth]{fig/motiv2lab} &
\includegraphics[width=0.16\textwidth]{fig/motiv3} &
\hbox{
\vbox{
\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-1} }
\hbox{\includegraphics[width=0.25\textwidth]{fig/ta-433}}
} 
}
\\
Protein 1CQW & Active site & Detected tunnels & Example of a  \\
             &            & (orange)          & trajectory
\end{tabular}
}
\caption{\label{fig::motiv}
    Tunnels in haloalkane dehalogenase with a possible trajectory of 1-chlorpropan.
}
\end{figure}

Understanding of interactions between proteins and other small molecules is crucial in many research fields, including drug design and protein engineering. 
These interactions are highly influence by the ability to transport the small molecule, called ligand, to the protein active site.
The active site can be considered as a deeply buried inner cavity with the ability to interact with the incoming ligand.
To get the ligand to the active site, there has to be the transportation path connecting the protein outer environment with the active site.
This paths, called tunnels, have to be wide enough and the physico-chemical properties of the surrounding amino acids have to be compatible with the ligand (Fig.~\ref{fig::motiv}).

@@

The tunnels can be detected using Voronoi diagrams~\cite{yaffe2008,caver3}.
%The widely used tools for tunnel detection are based on Voronoi diagrams~\cite{yaffe2008,caver3}.  %or a 3D grid~\cite{sehnal2013mole,petrek2006caver}.
The tunnels are computed assuming a spherical ligand (probe), and they are represented as a sequence of spheres.
Bio-chemists then decide if such a tunnel can be used to transport a given ligand mainly based on the tunnel lengths and bottleneck (i.e.,
        the radius of the smallest sphere that can traverse the whole tunnel).
Typical ligands are however non-spherical and it is therefore difficult to estimate their traversability in tunnels computed for a spherical probe.
The decision based only on the spherical tunnels requires previous expertise in the domain and yet it may be imprecise.

@@

@@

To guide tree growth of the tree through the tunnel, a moving virtual goal is used~\cite{vonasek2009rrt}.
The virtual goal $v, 1\le v \le n$, is the index of a tunnel sphere.
The random samples $\qrand$ are generated with probability $1-\gb$ from $\C$, otherwise they are generated
around the sphere $t_v \in T$.
%around which the random samples are generated with probability $\gb$.
%The random samples are not generated only from whole $\C$, as in the original RRT, but the sampling
%is guided using moving virtual goal~\cite{vonasek2009rrt}.
After the tree reaches the sphere $t_v$, i.e., the distance of the tree to the center of $t_v$ is
less than a predefined threshold $\dt$, the virtual goal is moved to successor of the last sphere in the tunnel
that is reached by the tree (lines~\ref{alg::main:a}--\ref{alg::main:b} in Alg.~\ref{alg::main}).
Setting the virtual goal to this successor allows the tree to avoid such parts of the tunnels that are
not traversable or reachable by the ligand.
This is necessary in the dense protein structures, where it is not always possible to exactly follow the tunnel.
The algorithm terminates after the predefined number of planning trials $\Imax$ or if the tree reaches
the last sphere in the tunnel, i.e., when $v = n$.

To generate the samples $\qrand$ around the virtual goal $v$, the translation part $(x,y,z)$ of $\qrand$ is generated
from $N(t_v,\Sigma)$, where $\Sigma$ is diagonal matrix with diagonal entries equal to a parameter $\rv$, and the rotational
part of $\qrand$ is generated using techniques described in~\cite{kuffnerES}.
The other two parameters (ligand index $l$ and scale $s$) of $\qrand$ can be left zero, as these are not used in the employed
metric for the nearest-neighbor search.

@@

The result of the proposed planning method is a tree of collision-free configurations, from which a path leading from $\qinit$
to the nearest node towards $\qgoal$ can be found.
Due to stochastic nature of RRT, it is not guaranteed that the tree can always reach the goal configuration $\qgoal$.
Despite, a trajectory leading as close as possible towards $\qgoal$ is extracted from the tree, as even these non-complete 
trajectories provide valuable information about the ligand behavior.

@@

The tested ligands are bigger than the tunnel with, therefore the ligands have been scaled down, and four
different scales were used: $\smin=\{0.3,0.4,0.5,0.6\}$.
No trajectories were found for $\smin > 0.6$.
For each of this scale, 50 collision-free initial states were found in the $2$\AA radius of the first sphere of the tunnel.
For each ligand, each minimal scale $\smin$ and each start configuration, 100 trajectories were computed using the proposed planner.
The parameters of the planner are: $\Imax=10000$, $m=50$, $\gb=0.9$, $\dt=1.5$\AA.
For each initial configuration, the success rate is computed as the ratio of trajectories that reached the end of the tunnel (to the distance $\rv$ or less) over the 100 trials.

The average runtimes and sizes of the build tree are shown in Tab.~\ref{tab::rrt}.
The tables also show the minimal, maximal success rate and the average success rate.
The average success rate is compute as the average value of the success rates achived for all 50 initial configurations.

The highest average success rate is achived for the most scaled-down ligands ($\smin=0.3$) and it decreases with the increasing $\smin$. 
The minimal and maximal success rates indicate how the initial position of the ligand influence the ability to pass the tunnels.
The ligand $\LB$, which was represented by more conformations than $\LA$, has higher minimal and also average success rate, which
indicates that the flexibility is important for the tunnel traversability.
The runtime is significantly higher for the $\LB$ ligand, which is caused by the larger number of conformations that need to be
examined in each expansion.

@@

\begin{table}
\centering
\caption{\label{tab::rrt}
    Runtime and sucess ratio of the planner
}
{
\renewcommand{\tabcolsep}{1pt}
\input table.003.tex
\hskip 5pt   
\input table.004.tex
}
\end{table}

@@

\section{Discussion \& implementation details}

Sampling-based motion planning is a powerful tool for computing tajectories for flexible ligands.
The proposed extensions to the RRT planner have been motivated by the necessity to cope with the narrow passage
and the high dimensionaly of the flexible ligands.
By scaling-down the geometry of ligand, which is also used in other related tools~\cite{cortes2005path}, 
the relative volume of the free region of the configuration space increases, which increases probability of placing
random samples $\qrand$ into the free region.
To further cope with the narrow passage problem, the translation part of the random samples is generated using
the 3D centerline of the tunnel.
The sampling along the tunnel centerline (in the ditance defined by $\rv$) does not only lead to better growth of the tree in the tunnel, but it also limits the maximal deflection of the trajectory from the tunnel.
Therefore, the bounding box of $\C$ can be defined around the tunnel in the distance $\rv$, which is typically
smaller than the bounding box of the whole protein.
This, together with the scaled-down ligand, increases relative volume of the free configuration space and help the RRT to
reach the goal configuration.

The runtime of the planner is curently given mainly by the collision-detection.
As the ligand is assumed to travel only along the tunnel centerline (in the distance defined by $\rv$),
only atoms located in the vicinity of the tunnel can be considered for collision detection.
A fast collision-detection can be achieved either using fast hiearchical data structures as OBB (implemented e.g. in~\cite{ozcollide})
    or using collision-detection system developed directly for proteins~\cite{ruiz2005biocd}.
